Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-23 paper.tex
Pass through results benchmarks

To demonstrate the accessibility of genotype data and efficiency with 
which calculations can be performed under the different formats,
we use the \texttt{bcftools +af-dist} plugin as an example.
% The details of the \texttt{af-dist} operation are not important:
% as an example of a whole-matrix operation. 
We chose this particular operation for several reasons. 
First, it is a straightforward calculation that 
requires examining every element in the genotype matrix,
and can be reproduced in different programming languages
without too much effort.
Secondly, it produces a small volume of output (a table of
deviations from Hardy-Weinberg expectations in ten allele frequency
bins) and therefore the time spent outputting results is negligible.
Finally, it has an efficient implementation written using the 
\texttt{htslib} C API~\citep{bonfield2021htslib}, distributed as 
a \texttt{bcftools} plugin.

@@

Fig~\ref{fig-whole-matrix-compute} shows timing results 
for running \texttt{bcftools +af-dist} and equivalent operations 
on the data of Fig~\ref{fig-data-storage}. There is a large
difference in the time required (note the log-log scale). 
The slowest approach uses Genozip. Because genozip does not
provide an API and only outputs VCF text, the best approach available 
is to pipe its output into \texttt{bcftools +af-dist}. Thus,
not only must we decode the data from the genozip format, we must 
then parse the large volumes of VCF text produced and then finally do the 
actual calculation. 
Running \texttt{bcftools +af-dist} directly on the gzipped VCF
is substantially faster, indicating that Genozip's excellent
compression performance comes at a substantial decompression cost.
Using a BCF file is again significantly faster,
because the packed binary format avoids the overhead of parsing 
VCF text into \texttt{htslib}'s internal data structures. 
We only use BCF for subsequent \texttt{bcftools} benchmarks.

>>>>

To demonstrate the accessibility of genotype data and efficiency with 
which calculations can be performed under the different formats,
we use the \texttt{bcftools +af-dist} plugin
(which computes a table of
deviations from Hardy-Weinberg expectations in 
allele frequency bins) as an example.
% The details of the \texttt{af-dist} operation are not important:
% as an example of a whole-matrix operation. 
We chose this particular operation for several reasons. 
First, it is a straightforward calculation that 
requires examining every element in the genotype matrix,
and can be reproduced in different programming languages
without too much effort.
Secondly, it produces a small volume of output  and therefore the 
time spent outputting results is negligible.
Finally, it has an efficient implementation written using the 
\texttt{htslib} C API~\citep{bonfield2021htslib},
and therefore running this command on a VCF or BCF file provides 
a reasonable approximation of the limit of what can be achieved in terms 
of whole-matrix computation on these formats.

@@

Fig~\ref{fig-whole-matrix-compute} shows timing results 
for running \texttt{bcftools +af-dist} and equivalent operations 
on the data of Fig~\ref{fig-data-storage}. There is a large
difference in the time required (note the log-log scale). 
The slowest approach uses Genozip. Because Genozip does not
provide an API and only outputs VCF text, the best approach available 
is to pipe its output into \texttt{bcftools +af-dist}. 
This involves first decoding the data from Genozip format,
then generating large volumes  of VCF text (terabytes, in the 
largest examples here), which we must 
subsequently parse before finally doing the actual calculation. 
Running \texttt{bcftools +af-dist} directly on the gzipped VCF
is substantially faster, indicating that Genozip's excellent
compression performance comes at a substantial decompression cost.
Using a BCF file is again significantly faster,
because the packed binary format avoids the overhead of parsing 
VCF text into \texttt{htslib}'s internal data structures. 
We only use BCF for subsequent \texttt{bcftools} benchmarks.
